---
title: "68806_EX2_PMmodel"
author: "YBD"
date: "2016 M11 15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## ET Estimation using Penman-Montieth Equation

This file shows the solution to Ex.2 for the P-M model of ET estimation and some of its adjustments

```{r metdata ,echo=FALSE}
metdata = read.table("metdata.txt", sep="", header=T)
summary(metdata)
```

# calculate daylight hours in Jerusalem to be used as input to P-M function

```{r , echo=FALSE}
# estimate daylength in Jerusalem
# day length in Jerusalem ranges between 10 to 14 h
# estimate day length for each half of the year and concatanate
daylength<-seq(from = 10, to = 14, length.out =((365-1)/2))
daylength2<-seq(from = 14, to = 10, length.out =((365)/2))
daylengthJer<-c(daylength,daylength2)

#fit the right daylength according to the day of the year
daynum<-metdata$yd
daylengthbynum=rep(0,length(daynum))

for (ii in 1:length(daylengthbynum)) {
  daylengthbynum[ii]=daylengthJer[daynum[ii]]
}

plot(daylengthJer, type='l',main='day length',xlab='day of the year',ylab='day length [h]')
```

calculate ET using the Penamn-Montieth equation of Naomi

```{r , echo=TRUE}
#' Penman Montieth Equation
#' 
#' THis function computer evapotranspiration based on radiation, conductance etc
#' 

#' @param        Tair    (deg C) air temperature
#' @param    vpd     (Pa)    vapour pressure deficit
#' @param     Rnet    (j/m2/day)      net radiation
#' @param     gs      (s/mm)  surface conductance
#' @param     ga      (s/mm)  aerodynamic conductance
#' @param     CP  =      1010.0 (J/kg*K) specific heat of air
#' @param    Pair = 101325 (Pa)
#' @author Naomi
#' @return Evapotranspiration (mm/day)
 
 
penman_montieth =
function(Tair, vpd, Rnet, gs,ga, dayl, CP=1010, Pair=101325) {
  
  #       Internal Variables
  #
  #       rho     (kg/m3)         density of air
  #       CP      (K/jg/degC)     specific heat of air
  #       lhvap   (J/kg)          latent heat of vapourization H20
  #       s       (Pa/degC)       slope of sat vpd vs T curve
  #       rs      (s/m)   surface resistance
  #       ra      (s/m)   aerodynamic resistance
  

  # convert Rnet to daytime value in j/m2/s
  Rnet = Rnet / (60*60*dayl)
  
  # convert conductance to resistance and change units
  rs = 1000.0/gs
  ra = 1000.0/ga
  
  # Assign tk (Kelvins)
  tk = Tair + 273.15
  
  #       Density of air (rho) as a fn. of air temp.
  rho = 1.292 * ( 0.00428 * Tair )
  
  #       Latent heat of vapourization as a fn. of Tair.
  lhvap = 2.5023e6 - 2430.54 * Tair
  
  #       Temperature offsets for slope estimates
  dt = 0.2
  t1 = Tair + dt
  t2 = Tair - dt
  
  #       Saturation vapour pressures at t1 and t2(Pa)
  pvs1 = 610.7 * exp(17.38 * t1 / ( 239.0 + t1))
  pvs2 = 610.7 * exp(17.38 * t2 / ( 239.0 + t2))
  
  #       Slope of pvs vs T curve at Tair (Pa/deg C)
  s = ( pvs1 -   pvs2 ) / ( t1 -  t2 )
  
  #       Calculate gamma
  gamma = CP * Pair / ( lhvap )
  
  
  #       Evaporation in W/m2
  et = ((s*Rnet) + (rho*CP*vpd/ra)) / (gamma*(1.0 + rs/ra) +s)
  
  
  #       mH20/s = W/m2 * 1kgH20/lhvap J * 1m3H20/1000kGH20
  ewater = ( et/ ( lhvap * 1000 ))
  
  # mmH20/day
  ewater.day = ewater * dayl*60*60 * 1000
  
  # return from your function
  ewater.day
}

```

```{r , echo=FALSE}

# calculate et using the PM function of N.Tague
et<-penman_montieth (Tair=metdata$tavg,vpd = metdata$vpd, Rnet=metdata$rnet,gs=metdata$gs, ga=metdata$ga, dayl=daylengthbynum, CP=1010, Pair=101325)
plot(et,type='l',main='ET P-M estimation', xlab='day', ylab='ET')

```

Adjust gs values following simplistic assumptions in Naomi email
if the temperature is above mean temperature, the maximum gs value is assigend
if the temperature is below mean temperature, a zero value is assigend

```{r , echo=TRUE}
# this function modifies the gs value according to the Naomi's example

#' @param        Tair    (deg C) air temperature
#' @param     gs      (s/mm)  surface conductance
#' @author Yoav BD
#' @return new gs (s/mm)

modified_gs=
  function(gs, Tair){
  cond1=mean(Tair)
  maxgs=max(gs)
  # calculate ET
  newgs=ifelse (Tair >= cond1, maxgs, 0)
}
```

```{r , echo=FALSE}
# evaluate a new gs using a basic if function
new_gs <-modified_gs(gs=metdata$gs, Tair=metdata$tavg)
et_newgs<-penman_montieth (Tair=metdata$tavg,vpd = metdata$vpd, Rnet=metdata$rnet,gs=new_gs, ga=metdata$ga, dayl=daylengthbynum, CP=1010, Pair=101325)

# plot both et calculated values
plot(et, main='comparison of original and modified ET',col="blue",ylab='ET', xlab='day')
points(et_newgs, main='comparison of original and modified ET',col="red")
# plot a comparison of the two et values
plot(et,et_newgs,type="p",col="red", main='comparison of original and modified ET',xlab='PM derived ET', ylab='modified ET using new gs value')
```

# adjust gs according to a second set of assumptions
gs is a multiple of maximum gs value multiplied by a vpd scaling factor and a linear gs-t relationship
```{r , echo=TRUE}
# Modify gs values:
# gs is defined by its max value, multiplied with relative Tair scale factor and scaled vpd

#' @param    Tair    (deg C) air temperature
#' @param    vpd     (Pa)    vapour pressure deficit
#' @param    gs      (s/mm)  surface conductance
#' @author Yoav BD
#' @return Evapotranspiration (mm/day)


modified_gs2 =
  function(Tair, vpd, gs) {
    # find max value of gs 
    maxgs=max(gs)

    # parametrize a linear gs-t relationship
    xval=seq(from=min(Tair) , to=max(Tair), by=0.001 )
    yval=seq(from=0.8 , to=1, length.out =length(xval))
    gstfactor=rep(0,length(gs))
    for (ii in 1:length(gs)) {
      gstfactor[ii]=yval[match(Tair[ii],xval)]
    }

    # scale the vpd values
    scaledvpd=vpd/max(vpd)
    
    newgs=maxgs*scaledvpd*gstfactor
  }
```

```{r , echo=FALSE}
new_gs2 <-modified_gs2(gs=metdata$gs, Tair=metdata$tavg, vpd=metdata$vpd)
et_newgs2<-penman_montieth (Tair=metdata$tavg,vpd = metdata$vpd, Rnet=metdata$rnet,gs=new_gs2, ga=metdata$ga, dayl=daylengthbynum, CP=1010, Pair=101325)

# plot both et calculated values
plot(et,type='l', main='comparison of original and modified ET',col="blue",ylab='ET', xlab='day')
lines(et_newgs, col="red")
lines(et_newgs2, col="green")

# plot a comparison of the two et values
plot(et,et_newgs2,type="p",col="red", main='comparison of original and modified ET',xlab='PM derived ET', ylab='modified ET using new gs value')
```
