---
title: "68806_EX2_PMmodel"
author: "YBD"
date: "2016 M11 15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## ET Estimation using Penman-Montieth Equation

This file shows the solution to Ex.2 for the P-M model of ET estimation and some of its adjustments

```{r metdata ,echo=FALSE}
metdata = read.table("metdata.txt", sep="", header=T)
summary(metdata)
```

# calculate daylight hours in Jerusalem to be used as input to P-M function

```{r , echo=FALSE}
# estimate daylength in Jerusalem
# day length in Jerusalem ranges between 10 to 14 h
# estimate day length for each half of the year and concatanate
daylength<-seq(from = 10, to = 14, length.out =((365-1)/2))
daylength2<-seq(from = 14, to = 10, length.out =((365)/2))
daylengthJer<-c(daylength,daylength2)

#fit the right daylength according to the day of the year
daynum<-metdata$yd
daylengthbynum=rep(0,length(daynum))

for (ii in 1:length(daylengthbynum)) {
  daylengthbynum[ii]=daylengthJer[daynum[ii]]
}

#add day hours to data matrix
metdata=cbind(metdata,daylengthbynum)


plot(daylengthJer, type='l',main='day length',xlab='day of the year',ylab='day length [h]')
```

calculate ET using the Penman-Montieth equation of Naomi

```{r , echo=TRUE}
#' Penman Montieth Equation
#' 
#' THis function computer evapotranspiration based on radiation, conductance etc
#' 

#' @param        Tair    (deg C) air temperature
#' @param    vpd     (Pa)    vapour pressure deficit
#' @param     Rnet    (j/m2/day)      net radiation
#' @param     gs      (s/mm)  surface conductance
#' @param     ga      (s/mm)  aerodynamic conductance
#' @param     CP  =      1010.0 (J/kg*K) specific heat of air
#' @param    Pair = 101325 (Pa)
#' @author Naomi
#' @return Evapotranspiration (mm/day)
 
 
penman_montieth =
function(Tair, vpd, Rnet, gs,ga, dayl, CP=1010, Pair=101325) {
  
  #       Internal Variables
  #
  #       rho     (kg/m3)         density of air
  #       CP      (K/jg/degC)     specific heat of air
  #       lhvap   (J/kg)          latent heat of vapourization H20
  #       s       (Pa/degC)       slope of sat vpd vs T curve
  #       rs      (s/m)   surface resistance
  #       ra      (s/m)   aerodynamic resistance
  

  # convert Rnet to daytime value in j/m2/s
  Rnet = Rnet / (60*60*dayl)
  
  # convert conductance to resistance and change units
  rs = 1000.0/gs
  ra = 1000.0/ga
  
  # Assign tk (Kelvins)
  tk = Tair + 273.15
  
  #       Density of air (rho) as a fn. of air temp.
  rho = 1.292 * ( 0.00428 * Tair )
  
  #       Latent heat of vapourization as a fn. of Tair.
  lhvap = 2.5023e6 - 2430.54 * Tair
  
  #       Temperature offsets for slope estimates
  dt = 0.2
  t1 = Tair + dt
  t2 = Tair - dt
  
  #       Saturation vapour pressures at t1 and t2(Pa)
  pvs1 = 610.7 * exp(17.38 * t1 / ( 239.0 + t1))
  pvs2 = 610.7 * exp(17.38 * t2 / ( 239.0 + t2))
  
  #       Slope of pvs vs T curve at Tair (Pa/deg C)
  s = ( pvs1 -   pvs2 ) / ( t1 -  t2 )
  
  #       Calculate gamma
  gamma = CP * Pair / ( lhvap )
  
  
  #       Evaporation in W/m2
  et = ((s*Rnet) + (rho*CP*vpd/ra)) / (gamma*(1.0 + rs/ra) +s)
  
  
  #       mH20/s = W/m2 * 1kgH20/lhvap J * 1m3H20/1000kGH20
  ewater = ( et/ ( lhvap * 1000 ))
  
  # mmH20/day
  ewater.day = ewater * dayl*60*60 * 1000
  
  # return from your function
  ewater.day
}

```

```{r , echo=FALSE}

# calculate et using the PM function of N.Tague
et<-penman_montieth (Tair=metdata$tavg,vpd = metdata$vpd, Rnet=metdata$rnet,gs=metdata$gs, ga=metdata$ga, dayl=metdata$daylengthbynum, CP=1010, Pair=101325)
plot(et,type='l',main='ET P-M estimation', xlab='day', ylab='ET')

```

Adjust gs values following simplistic assumptions in Naomi email
if the temperature is above mean temperature, the maximum gs value is assigned
if the temperature is below mean temperature, a zero value is assigned

```{r , echo=TRUE}
# this function modifies the gs value according to the Naomi's example

#' @param        Tair    (deg C) air temperature
#' @param     gs      (s/mm)  surface conductance
#' @author Yoav BD
#' @return new gs (s/mm)

modified_gs=
  function(gs, Tair){
  cond1=mean(Tair)
  maxgs=max(gs)
  # calculate ET
  newgs=ifelse (Tair >= cond1, maxgs, 0)
}
```

```{r , echo=FALSE}
# evaluate a new gs using a basic if function
new_gs <-modified_gs(gs=metdata$gs, Tair=metdata$tavg)
et_newgs<-penman_montieth (Tair=metdata$tavg,vpd = metdata$vpd, Rnet=metdata$rnet,gs=new_gs, ga=metdata$ga, dayl=metdata$daylengthbynum, CP=1010, Pair=101325)

# plot both et calculated values
plot(et, main='comparison of original and modified ET',col="blue",ylab='ET', xlab='day')
points(et_newgs, main='comparison of original and modified ET',col="red")
# plot a comparison of the two et values
plot(et,et_newgs,type="p",col="red", main='comparison of original and modified ET',xlab='PM derived ET', ylab='modified ET using new gs value')
```

# adjust gs according to a second set of assumptions
gs is a multiple of maximum gs value multiplied by a vpd scaling factor and a linear gs-t relationship
```{r , echo=TRUE}
# Modify gs values:
# gs is defined by its max value, multiplied with relative Tair scale factor and scaled vpd

#' @param    Tair    (deg C) air temperature
#' @param    vpd     (Pa)    vapour pressure deficit
#' @param    gs      (s/mm)  surface conductance
#' @author Yoav BD
#' @return Evapotranspiration (mm/day)


modified_gs2 =
  function(Tair, vpd, gs) {
    # find max value of gs 
    maxgs=max(gs)

    # parametrize a linear gs-t relationship
    xval=seq(from=min(Tair) , to=max(Tair), by=0.001 )
    yval=seq(from=0.8 , to=1, length.out =length(xval))
    gstfactor=rep(0,length(gs))
    for (ii in 1:length(gs)) {
      gstfactor[ii]=yval[match(Tair[ii],xval)]
    }

    # scale the vpd values
    scaledvpd=vpd/max(vpd)
    
    newgs=maxgs*scaledvpd*gstfactor
  }
```

```{r , echo=FALSE}
new_gs2 <-modified_gs2(gs=metdata$gs, Tair=metdata$tavg, vpd=metdata$vpd)
et_newgs2<-penman_montieth (Tair=metdata$tavg,vpd = metdata$vpd, Rnet=metdata$rnet,gs=new_gs2, ga=metdata$ga, dayl=metdata$daylengthbynum, CP=1010, Pair=101325)

# plot both et calculated values
plot(et,type='l', main='comparison of original and modified ET',col="blue",ylab='ET', xlab='day')
lines(et_newgs, col="red")
lines(et_newgs2, col="green")

# plot a comparison of the two et values
plot(et,et_newgs2,type="p",col="red", main='comparison of original and modified ET',xlab='PM derived ET', ylab='modified ET using new gs value')
```

# sensitivity estimation
visually estimate the effect of several parameters of the PM function
each variable is separately multiplied by a factor running from 0 to 2 with differences of 0.05
blue lines indicate factors smaller than 1, red lines indicate factor larger than 1, and black line indicates original estimation of ET values
```{r , echo=FALSE}
senseitivityfactor=seq(from=0.00 ,to=2.0 ,by=0.05)
et_sensitivity=array(data=0,dim=c(length(et),length(senseitivityfactor),6))

# define the colums to be evaluated for sensitivity
sensitivity_vec=c(8,9,10,11,12,17)
jj=1;

for (jj in 1:length(sensitivity_vec)) {
  for (ii in 1:length(senseitivityfactor)){
    temp_metdata=metdata
    temp_metdata[,sensitivity_vec[jj]]=metdata[,sensitivity_vec[jj]]*senseitivityfactor[ii]
    et_sensitivity[1:length(et),ii,jj]<-penman_montieth (Tair=temp_metdata$tavg,vpd = temp_metdata$vpd, Rnet=temp_metdata$rnet,gs=temp_metdata$gs, ga=temp_metdata$ga, dayl=temp_metdata$daylengthbynum, CP=1010, Pair=101325)
  }
}

# jj defines the parameter to be plotted
jj=1;

#define a color vector for plotting data
colorvec<-c(rep('blue', 20), 'black',rep('red', 20))

plot(et,type="l",col="black", main='ET sensitivity to temperature',xlab='day', ylab='PM estimated ET')
for(ii in 1:length(senseitivityfactor)){
lines(et_sensitivity[,ii,jj],col=colorvec[ii])}
lines(et,type="l",col="black")

# jj defines the parameter to be plotted
jj=2;

#define a color vector for plotting data
colorvec<-c(rep('blue', 20), 'black',rep('red', 20))

plot(et,type="l",col="black", main='ET sensitivity to vpd',xlab='day', ylab='PM estimated ET')
for(ii in 1:length(senseitivityfactor)){
lines(et_sensitivity[,ii,jj],col=colorvec[ii])}
lines(et,type="l",col="black")

# jj defines the parameter to be plotted
jj=3;

#define a color vector for plotting data
colorvec<-c(rep('blue', 20), 'black',rep('red', 20))

plot(et,type="l",col="black", main='ET sensitivity to Rnet',xlab='day', ylab='PM estimated ET')
for(ii in 1:length(senseitivityfactor)){
lines(et_sensitivity[,ii,jj],col=colorvec[ii])}
lines(et,type="l",col="black")

# jj defines the parameter to be plotted
jj=4;

#define a color vector for plotting data
colorvec<-c(rep('blue', 20), 'black',rep('red', 20))

plot(et,type="l",col="black", main='ET sensitivity to gs',xlab='day', ylab='PM estimated ET')
for(ii in 1:length(senseitivityfactor)){
lines(et_sensitivity[,ii,jj],col=colorvec[ii])}
lines(et,type="l",col="black")


# jj defines the parameter to be plotted
jj=5;

#define a color vector for plotting data
colorvec<-c(rep('blue', 20), 'black',rep('red', 20))

plot(et,type="l",col="black", main='ET sensitivity to ga',xlab='day', ylab='PM estimated ET')
for(ii in 1:length(senseitivityfactor)){
lines(et_sensitivity[,ii,jj],col=colorvec[ii])}
lines(et,type="l",col="black")

# jj defines the parameter to be plotted
jj=6;
#define a color vector for plotting data
colorvec<-c(rep('blue', 20), 'black',rep('red', 20))

plot(et,type="l",col="black", main='ET sensitivity to day length',xlab='day', ylab='PM estimated ET')
for(ii in 1:length(senseitivityfactor)){
lines(et_sensitivity[,ii,jj],col=colorvec[ii])}
lines(et,type="l",col="black")
```

# Conclusions
the model appears to show high sensitivity to some of the inputs, but also demonstrates high resiliency to others
vpd and day length have little effect on the PM estimation of ET
Radiation and temperature show relatively high sensitivity, while gs shows relatively moderate impact
ga shows inverse sensitivity, and increasing factors produce smaller ET values, highest values are obtained for the smallest factors, while increasing ga shows moderate impact on ET. this relationship is easily explained by the location of the ga term in the PM equation